Locality and Effect Reflection

### 2025-07-09 paper.tex
Write additional things

  Effect reflection lets you work with any algebraic monad using an
  ordinary function type, allowing effectful code to be written in
  direct style. This provides all the benefits of typed algebraic
  effects and handlers without requiring an effect system. We
  demonstrate effect reflection by implementing it as a library in
  OxCaml\cite{lorenzen2024oxidizing,} and working through some classic
  examples of programming with effects.
\end{abstract}

@@

Note that, in OCaml, nothing prevents attempting to perform an algebraic
effect operation that is not supported by the current ambient monad,
resulting in an \lstinline[style=oxcaml]{Unhandled} exception being
raised. Other languages, e.g. Koka\cite{??}, use an effect system to
statically ensure operations are only performed when they are available.

@@

Treating the ambient monad as a notion of location allows a language to
provide an isomorphism:
\begin{equation*}
  \mathop{M}(\glob X) \qquad \cong \qquad Y \mathbin{\rightarrow} \glob X
\end{equation*}
where $M$ is any algebraic monad, $Y$ is a specific type related to that
monad, and $\glob$ is a comonadic global modality. We call this
isomorphism \emph{effect reflection}.

\todo{Restrict the above claim to signatures}

The ability to use the ordinary arrow type
$Y \mathbin{\rightarrow} \glob X$ in place of the monad
$\mathop{M}(\glob X)$ allows us to conveniently write effectful
code in direct style, just as algebraic effects do.

@@

Effect reflection is closely related to \emph{lexical effect handlers} in
systems such as Effeckt\cite{??}. Lexical effect handlers have a value that
represents the effect handler, but these values are second class and
require their own special typing rules.

Effect reflection is also closely related to \emph{named effect
  handlers}\cite{??}. Named effect handlers use higher-rank polymorphism
to emulate locality tracking in much the same way as the
\lstinline[style=haskell]{St} monad in Haskell. Using higher-rank
polymorphism prevents global type inference, requiring complex inference
that extends beyond classic Hindly-Milner, and demanding a higher
annotation burden on the programmer. \todo{Mention the explicit monad
that named effect handlers presumably need.}

Effect reflection is a variation on monadic reflection\cite{??}.
\todo{Make this relationship precise.}

@@

These presheaf semantics are closely related to presheaf semantics for
higher-order abstract syntax\cite{??}.

@@

  val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
end
\end{lstlisting}
Using \lstinline[style=oxcaml]{perform} we can implement reflection of
terms as:
\begin{lstlisting}[style=oxcaml]
let rec reflect t h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect (k (perform op h)) h

let rec reflect_local (t @ local) h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect_local (k (perform op h)) h
\end{lstlisting}
with types:
\begin{lstlisting}[style=oxcaml]
val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)

val reflect_local :
   'a Term(O).t @ local -> (Handler(O).t @ local -> 'a) @ local
\end{lstlisting}

@@

This ability for higher-order functions like
\lstinline[style=oxcaml]{List.iter} to be reused with different ambient
monads achieves the ``lightweight effect polymorphism'' described by ??
et al.\cite{??} without requiring a separate notion of second-class
function.

@@

\subsection{Effect sums}

@@

\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}.

Terms in the free algebras of the sum of two effects
(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other\cite{??}. We can definet this type as:
\begin{lstlisting}[style=oxcaml]
module Half_term (L : Op) (R : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
end
\end{lstlisting}
and expose the isomorphism between
\lstinline[style=oxcaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
and \lstinline[style=oxcaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
\begin{lstlisting}[style=oxcaml]
module Half_reflection (L : Op) (R : Op) : sig
  module L_eff = Reflection(L)
  module R_eff = Reflection(R)
  module Sum_eff = Reflection(Sum(L)(R))
  module Proj = Project(L)(R)

  let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
    let rec loop lh = function
      | Term(Sum(L)(R)).Return ->
          Half_term(L)(R).Return v
      | Term(Sum(L)(R)).Op(Left op, k) ->
          loop lh (k (L_eff.perform op lh))
      | Term(Sum(L)(R)).Op(Right op, k) ->
          Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
    in
    loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh

  let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
    match f lh with
    | Return x -> x
    | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
end
\end{lstlisting}

@@

type aints =
  | Finished
  | More of int * (Handler(Await).t @ local -> ints)

let handle_gen_in_await f ah =
    let rec loop = function
    | Half_term(Await)(Gen).Return () -> Finished
    | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, k ())
  in
  loop (Await_gen_eff.reify f)
\end{lstlisting}
where \lstinline[style=oxcaml]{handle_gen_in_await} has the type:
\begin{lstlisting}[style=oxcaml]
val handle_gen_in_await :
  (Handler(Await).t @ local -> Handler(Gen).t @ local -> 'a)
  -> Handler(Await).t @ local
  -> aints
\end{lstlisting}

@@

We define the operational semantics as a pair of mutually defined
reduction relations $\reducestov$ and $\reducestoc$ for values and
computations respectively shown in Fig.\ref{fig:reduction} The
definitions of normal forms for values and computations are given in
Fig.\ref{fig:normal}. Subject reduction and strong normalization of
these rules can be proved using standard techniques.

@@

Note that the reduction relations do not include $\eta$ expansion. They
also do not include some conversions you might expect for computations:
\begin{itemize}
\item There is no rule to convert
  $\; \letv{x_2}{\letv{x_1}{c_1}}{c_2}{c_3} \;$ to
  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \;$ or vice versa.
\item There is no rule to convert $\; \letv{x}{c}{\return x} \;$ to $c$.
\end{itemize}
These rules are excluded because they are not $\beta$ rules. Instead
they are better thought of as $\nu$ rules \cite{??}: additional
equalities between neutral terms that are immediate consequences of
structural induction. Rather than be included as part of ordinary
reduction they can be done as an additional standardization pass on
neutral terms after reduction and $\eta$-expansion have completed. This
staged approach to computing equality of terms works because $\nu$ rules
cannot produce additional $\beta$ redexes.

@@

Using the global modality and $\grun{c}$ construct we can embed the
Simply Typed Lambda Calculus (STLC) into this language as values. We
define an embedding $\embed{\_}$ on types, contexts and terms of STLC:
\begin{equation*}
\begin{aligned}[c]
  &\embed{1} = 1 \\
  &\embed{A \times B} = \embed{A} \times \embed{B} \\
  &\embed{A \rightarrow B} = (\glob \embed{A}) \rightarrow \embed{B} \\ \\
  &\embed{\cempty} = \cempty \\
  &\embed{\lbind{\Gamma}{x}{A}} = \gbind{\embed{\Gamma}}{x}{\embed{A}}
\end{aligned}
\qquad
\begin{aligned}[c]
  &\embed{x} = x \\
  &\embed{(e, f)} = (\embed{e}, \embed{f}) \\
  &\embed{\fst{e}} = \fst{\embed{e}} \\
  &\embed{\snd{e}} = \snd{\embed{e}} \\
  &\embed{\abs{x}{A}{e}} =
    \abs{x^{\prime}}{\glob \embed{A}}{\return{(\gunbox{x}{x^{\prime}}{\embed{e}})}} \\
  &\embed{\app{f}{e}} = \grun{\app{\embed{f}}{(\gbox{\embed{e}})}}
\end{aligned}
\end{equation*}

Note that $\reducestov$ matches the behaviour of $\beta$ reduction of
STLC too, with each STLC reduction step corresponding to up to 3 steps
of $\reducestov$.

@@

\subsection{Effect sums and inductive types}

@@

The key to this process is the following lemma
\cite{hyland2006combining}.
\begin{lemma}
\label{lemma:sums}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(S) \cong \mu X.T_L(S + F_\Sigma(X))
  \end{equation*}
\end{lemma}
where $\mu X.F(X)$ is the intial algebra of a suitable endofunctor $F$,
and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$ is
the free monad.

Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
user via an inductive type. In particular, given $\Sigma$ and a
presheave $P$, we can define the following initial algebra:
\begin{align*}
&\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
&\ind{\Sigma}(P) = \mu X. (P + F_\Sigma \circ \mon{X})
\end{align*}
and observe the following variation on Lemma \ref{lemma:sums}.
\begin{equation*}
T_{L + \Sigma}(P(L)) \cong \mon{(\ind{\Sigma}(P))}(L)
\end{equation*}

@@

\subsection{Effect reification}

Exponentials in categories of presheaves are defined such that:
\begin{equation*}
\arrow{P}{Q}(\catobj) \cong \homset{\psh}{\yoneda{\catobj} \times P}{Q}
\end{equation*}
Applying this to $\arrow{\yoneda{L}}{\mon{(\glob P)}}$ and then applying
the Yoneda Lemma we get:
\begin{align*}
\arrow{\yoneda{L}}{\mon{(\glob P)}}(L')
  &\cong \homset{\pshlawcop}{\yoneda{L'} \times \yoneda{L}}{\mon{(\glob P)}} \\
  &\cong \homset{\pshlawcop}{\yoneda{L' + L}}{\mon{(\glob P)}} \\
  &\cong \mon{(\glob P)}(L' + L) \\
  &\cong T_{(L' + L)}(P(L'))
\end{align*}

In the case where $L$ is a signature $\Sigma$, we can further apply our
variation of Lemma \ref{lemma:sums}, and observing that all the steps in
this proof were functorial, produce this isomorphism:
\begin{equation*}
  \arrow{\yoneda{\Sigma}}{\mon{(\glob P)}} \cong \mon{(\ind{\Sigma}(\glob P))}
\end{equation*}
which we call $\reifyc{\Sigma}{}$.

$\reifyc{\Sigma}{}$ directly gives us our semantics for reification.
\begin{align*}
&\sem{\Gamma \turnc \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)} =
  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}) \\
\end{align*}

@@

\subsection{Implementation with untyped effect handlers}

We can implement the interface from Section \ref{sec:interface} directly
in terms of OCaml's untyped effect
handlers\cite{Sivaramakrishnan2021retrofitting}. The trick is to
generate a fresh effect constructor for each reification, pass that
constructor around as part of the \lstinline[style=oxcaml]{handler}, and
then use it to perform operations during reflection.

\begin{lstlisting}[style=oxcaml]
module Handler (O : Op) = struct
  type t = { perform : 'r. 'r O.t -> 'r }
end

@@

  let reflect t =
    fun {perform} ->
      let rec loop = function
        | Return v -> v
        | Op(o, f) -> loop (f (perform o))
      in
      loop t

@@

  let reflect_local t = exclave
    fun {perform} ->
      let rec loop = function
        | Return v -> v
        | Op(o, f) -> loop (f (perform o))
      in
      loop t

end
\end{lstlisting}
Sums are then implemented by just lifting the injections on operations:
\begin{lstlisting}[style=oxcaml]
module Sum (L : Op) (R : Op) = struct

  module Op = struct
    type 'a t =
      | Left : 'a L.t -> 'a t
      | Right : 'a R.t -> 'a t
  end

  let outl {perform} =
    {perform = fun op -> perform (Left op)}

  let outr {perform} =
    {perform = fun op -> perform (Right op)}

end
\end{lstlisting}

@@

In the above implementation, whenever an operation is reflected we
perform a linear search up the stack of handlers until we find the
corresponding one. This search is not actually necessary...

@@

@@

\subsection{Semantics of algebraic effects}

@@

@@

\section{Future work}
\label{sec:future-work}

>>>>

  Effect reflection lets you work with any algebraic monad using an
  ordinary function type, allowing effectful code to be written in
  direct style. This provides all the benefits of typed algebraic
  effects and handlers without requiring an effect system or other
  specific additions to the type system. We demonstrate effect
  reflection by implementing it as a library in
  OxCaml\cite{lorenzen2024oxidizing} and working through some classic
  examples of programming with effects.
\end{abstract}

@@

Note that, in OCaml, the effects are \emph{untracked}: nothing prevents
attempting to perform an algebraic effect operation that is not
supported by the current ambient monad, resulting in an
\lstinline[style=oxcaml]{Unhandled} exception being raised.

Other languages track effects to ensure that operations are only
performed when they are supported. However, these require extending the
type system with explicit mechanisms for tracking effects, as in
Koka\cite{leijen2017type}, or for tracking a general notion of
capability, as in Effekt\cite{brachthauser2022effects}.

The ``Lightweight effect polymorphism'' approach of Brachthauser et
al. elegantly manages to track effects without requiring explicit
effect-tracking machinery in the type system. However, it is restricted
to second-class functions. We would like to extend this approach to
first-class functions without having to fall back to explicit
effect-tracking machinery.

@@

Treating the ambient monad as a notion of location allows a language to
provide, for any algebraic signature $\Sigma$, an isomorphism:
\begin{equation*}
  \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
\end{equation*}
where $\ind{\Sigma}$ is an inductive type representing the free
algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
to inject terms of $\Sigma$ into the current ambient
monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
$\Sigma$. We call this isomorphism \emph{effect reflection}.

The ability to use the ordinary arrow type
$\yoneda{\Sigma} \mathbin{\rightarrow} \glob X$ in place of the monad
$\ind{\Sigma}(\glob X)$ allows us to conveniently write effectful code,
using multiple effects, in direct style, just as algebraic effects do.

@@

\todo{Add some mention of connection to lexical and named handlers}

@@

@@

  val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
end
\end{lstlisting}
Using \lstinline[style=oxcaml]{perform} we can implement reflection of
global terms as:
\begin{lstlisting}[style=oxcaml]
let rec reflect t h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect (k (perform op h)) h
\end{lstlisting}
with type:
\begin{lstlisting}[style=oxcaml]
val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
\end{lstlisting}
Reflection of local terms can be defined similarly.

@@

This ability for higher-order functions like
\lstinline[style=oxcaml]{List.iter} to be reused with different ambient
monads achieves the ``lightweight effect polymorphism'' described by
Brachth{\"a}user et al.\cite{brachthauser2020effects} without requiring
a separate notion of second-class function.

@@

\subsection{Effect sums and effect forwarding}

@@

\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}. Terms in
the free algebras of the sum of two effects
(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other\cite{hyland2006combining}. We can define
this type and associated isomorphism as:
\begin{lstlisting}[style=oxcaml]
module Half_term (L : Op) (R : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
end

module Half_reflection (L : Op) (R : Op) : sig
  val reify :
    (Handler(L).t @ local -> Handler(R).t @ local -> 'a)
    -> Handler(L).t @ local -> 'a Half_term(L)(R).t

  val reflect :
    (Handler(L).t @ local -> 'a Half_term(L)(R).t)
    -> Handler(L).t @ local -> Handler(R).t @ local -> 'a
end
\end{lstlisting}

Note that the key difference between
\lstinline[style=oxcaml]{Half_term(L)(R).t} and
\lstinline[style=oxcaml]{Term(R).t} is that continuing the computation
after an operation requires passing in a local
\lstinline[style=oxcaml]{Handler(L).t} again.

%\begin{lstlisting}[style=oxcaml]
%module Half_reflection (L : Op) (R : Op) : sig
%  module L_eff = Reflection(L)
%  module R_eff = Reflection(R)
%  module Sum_eff = Reflection(Sum(L)(R))
%  module Proj = Project(L)(R)
%
%  let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
%    let rec loop lh = function
%      | Term(Sum(L)(R)).Return ->
%          Half_term(L)(R).Return v
%      | Term(Sum(L)(R)).Op(Left op, k) ->
%          loop lh (k (L_eff.perform op lh))
%      | Term(Sum(L)(R)).Op(Right op, k) ->
%          Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
%    in
%    loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
%
%  let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
%    match f lh with
%    | Return x -> x
%    | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
%end
%\end{lstlisting}

@@

type aints =
  | Finished
  | More of int * (Handler(Await).t @ local -> aints)

let handle_gen_in_await f ah =
    let rec loop = function
    | Half_term(Await)(Gen).Return () -> Finished
    | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
  in
  loop (Await_gen_eff.reify f ah)
\end{lstlisting}
where \lstinline[style=oxcaml]{handle_gen_in_await} has the type:
\begin{lstlisting}[style=oxcaml]
val handle_gen_in_await :
  (Handler(Await).t @ local -> Handler(Gen).t @ local -> 'a)
  -> Handler(Await).t @ local
  -> aints
\end{lstlisting}

@@

We define the operational semantics as a pair of mutually defined
reduction relations $\reducestov$ and $\reducestoc$ for values and
computations respectively shown in Fig.\ref{fig:reduction}. The
definitions of normal forms for values and computations are given in
Fig.\ref{fig:normal}. Subject reduction and strong normalization of
these rules can be proved using standard techniques.

@@

Note that the reduction relations do not include $\eta$ expansion. They
also do not include some conversions you might expect for computations:
\begin{itemize}
\item There is no rule to convert
  $\; \letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} \,$ to
  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \,$ or vice versa.
\item There is no rule to convert $\; \letv{x}{c}{\return x} \,$ to $c$.
\end{itemize}
These rules are excluded because they are not $\beta$ rules. Instead
they are better thought of as $\nu$ rules \cite{??}: additional
equalities between neutral terms that are immediate consequences of
structural induction. Rather than be included as part of ordinary
reduction they can be done as an additional standardization pass on
neutral terms after reduction and $\eta$-expansion have completed. This
staged approach to computing equality of terms works because $\nu$ rules
cannot produce additional $\beta$ redexes.

@@

Using the global modality and $\grun{c}$ construct we can embed the
Simply Typed Lambda Calculus (STLC) into this language as values. We
define an embedding $\embed{\_}$ on types, contexts and terms of STLC in
Fig.\ref{fig:stlc}. Note that $\reducestov$ matches the behaviour of
$\beta$ reduction of STLC too, with each STLC reduction step
corresponding to up to 3 steps of $\reducestov$.
\begin{figure}
\begin{equation*}
\begin{aligned}[c]
  &\embed{1} = 1 \\
  &\embed{A \times B} = \embed{A} \times \embed{B} \\
  &\embed{A \rightarrow B} = (\glob \embed{A}) \rightarrow \embed{B} \\ \\
  &\embed{\cempty} = \cempty \\
  &\embed{\lbind{\Gamma}{x}{A}} = \gbind{\embed{\Gamma}}{x}{\embed{A}}
\end{aligned}
\qquad
\begin{aligned}[c]
  &\embed{x} = x \\
  &\embed{(e, f)} = (\embed{e}, \embed{f}) \\
  &\embed{\fst{e}} = \fst{\embed{e}} \\
  &\embed{\snd{e}} = \snd{\embed{e}} \\
  &\embed{\abs{x}{A}{e}} =
    \abs{x^{\prime}}{\glob \embed{A}}{\return{(\gunbox{x}{x^{\prime}}{\embed{e}})}} \\
  &\embed{\app{f}{e}} = \grun{\app{\embed{f}}{(\gbox{\embed{e}})}}
\end{aligned}
\end{equation*}
\caption{Embedding of Simply-Typed Lambda Calculus into values}
\label{fig:stlc}
\end{figure}

@@

\subsection{Effect reification and inductive types}

@@

The key to this process is the following lemma
\cite{hyland2006combining}.
\begin{lemma}
\label{lemma:sums}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(S) \cong \mu X.T_L(S + F_\Sigma(X))
  \end{equation*}
\end{lemma}
where $\mu X.G(X)$ is the intial algebra of some polynomial endofunctor
$G$, and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$
is the free monad.

Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
user via an inductive type. Categories of presheaves have initial
algebras of polynomial endofunctors, i.e. W-types, so we can construct
the required initial algebra as a presheave. Given an algebraic
signature $\Sigma$ and a presheave $P$, we define the following family
of initial algebras:
\begin{align*}
&\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
&\ind{\Sigma}(P) = \mu X. (P + F_\Sigma \circ \mon{X})
\end{align*}
and observe the following variation on Lemma \ref{lemma:sums}.
\begin{lemma}
\label{lemma:sums2}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(P(L)) \cong \mon{(\ind{\Sigma}(P))}(L)
  \end{equation*}
\end{lemma}

@@

Exponentials in categories of presheaves are defined such that:
\begin{equation*}
\arrow{P}{Q}(\catobj) \cong \homset{\psh}{\yoneda{\catobj} \times P}{Q}
\end{equation*}
Applying this to $\arrow{\yoneda{L}}{\mon{(\glob P)}}$ and then applying
the Yoneda Lemma we get:
\begin{align*}
\arrow{\yoneda{L}}{\mon{(\glob P)}}(L^{\prime})
  &\cong \homset{\pshlawcop}{\yoneda{L^{\prime}} \times \yoneda{L}}{\mon{(\glob P)}} \\
  &\cong \homset{\pshlawcop}{\yoneda{L^{\prime} + L}}{\mon{(\glob P)}} \\
  &\cong \mon{(\glob P)}(L^{\prime} + L) \\
  &\cong T_{(L^{\prime} + L)}(P(L^{\prime}))
\end{align*}

In the case where $L$ is a signature $\Sigma$, we can further apply
Lemma \ref{lemma:sums2} and, observing that all the steps in this proof
were natural in $L^{\prime}$, produce this isomorphism:
\begin{equation*}
  \arrow{\yoneda{\Sigma}}{\mon{(\glob P)}} \cong \mon{(\ind{\Sigma}(\glob P))}
\end{equation*}
which we call $\reifyc{\Sigma}{}$.

$\reifyc{\Sigma}{}$ gives us our semantics for reification.
\begin{align*}
&\sem{\Gamma \turnc \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)} =
  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}) \\
\end{align*}

@@

\subsection{Implementation with untracked effect handlers}

We can implement the interface from Section \ref{sec:interface} directly
in terms of OCaml's untracked effect
handlers\cite{Sivaramakrishnan2021retrofitting}. The trick is to
generate a fresh effect constructor for each reification, pass that
constructor around as part of the \lstinline[style=oxcaml]{handler}, and
then use it to perform operations during reflection. The full implementation
is given in Fig.\ref{fig:implementation}.
\begin{figure}
\begin{lstlisting}[style=oxcaml]
module Handler (O : Op) = struct

  type t = { perform : 'r. 'r O.t -> 'r }

end

@@

@@

  let perform t =
   fun h -> h.perform t

end

module Project (L : Op) (R : Op) = struct

  let outl {perform} =
    {perform = fun op -> perform (Sum(L)(R).Left op)}

  let outr {perform} =
    {perform = fun op -> perform (Sum(L)(R).Right op)}

end
\end{lstlisting}
\caption{Implementation}
\label{fig:implementation}
\end{figure}

@@

In the implementation in Fig.\ref{fig:implementation}, whenever an
operation is reflected we perform a linear search up the stack of
handlers until we find the corresponding one. This search is not
actually necessary: there is a one-to-one relationship between the
handlers and the handler values used to reflect operations. By modifying
the OxCaml runtime to allow values on the locals stack that directly
point to the fibers used to implement effect handlers we can reflect
operations by jumping directly to the relevant handler. This is roughly
the implementation technique described by Ma et al.\cite{ma2024lexical}.

@@

Effect reflection is closely related to \emph{lexical effect handlers}
in systems such as Effekt\cite{brachthauser2020effects} or Biernacki et
al. \cite{biernacki2019binders}. Lexical effect handlers have a value
that represents the effect handler, similar to $\yoneda{\Sigma}$ values,
but these values are second class and require their own special typing
rules. More recent work on Effekt\cite{brachthauser2022effects}
addresses the second-class nature of these values, but requires further
extending the type system by tracking the set of capabilities used by
each value.

Effect reflection is also closely related to \emph{named effect
  handlers}\cite{xie2022first}. Named effect handlers use higher-rank
polymorphism to emulate locality tracking in much the same way as the
\lstinline[style=haskell]{St} monad in Haskell. This still requires
using an effect system to track which named effects are used by a
computation, in order to ensure that those effects do not escape the
scope of their handler. Using higher-rank polymorphism also prevents
global type inference, requiring complex inference that extends beyond
classic Hindly-Milner, and demanding a higher annotation burden on the
programmer.

@@

Effect reflection is a variation on monadic
reflection\cite{filinski1999representing}. By restricting ourselves to
algebraic monads with no equations we remove the need to manually define
\emph{glue} that describes the interaction between the monad being
reified and the current ambient monad. By using locality and
$\yoneda{\Sigma}$ values we avoid the need to have an effect system.

\subsection{Semantics of algebraic effect handlers}

Hyland et al.\cite{hyland2006combining} investigate combining algebraic
effects, including the key property of sums of algebraic theories (our
Lemma \ref{lemma:sums}) on which handlers are based.

The original work on algebraic effect
handlers\cite{plotkin2009handlers,plotkin2013handling} gives a semantics
in terms of free algebras but the semantics work with a fixed algebraic
theory: handlers do not actually change the ambient monad.

@@

Our use of a comonadic global modality to track values which do not
depend on the ambient monad is essentially the same as the purity
comonad described by Choudhury and
Krishnaswami\cite{choudhury2020recovering}. Their semantic description
of their comonad is very different -- based on a semantics of
capabilities -- but the typing rules, and the practical result, are the
same.

@@

Our presheaf semantics are closely related to presheaf semantics for
higher-order abstract syntax\cite{hu2022category}. In both cases
presheaves are used to index the semantics of types and terms by some
notion of context, with the Yoneda embedding used to access this context
and the global modality used to control it.
